


\def\codeTabSpace{\hspace*{6mm}}
\newenvironment{code}%
{\begin{tabbing}%
		\codeTabSpace \= \hspace*{70mm} \= \hspace*{42mm} \= \kill%
	}%
	{\end{tabbing}%
}
\newcounter{ind}
\newcommand{\n}{\addtocounter{ind}{7}\hspace*{7mm}}
\newcommand{\p}{\addtocounter{ind}{-7}\hspace*{-7mm}}
\newcommand{\nlc}{\\\stepcounter{linenum}{\scriptsize \arabic{linenum}}\>\hspace*{\value{ind}mm}}
\newcommand{\ul}{\\\>\hspace*{\value{ind}mm}}
\newcommand{\bl}{\\[-1.5mm]\>\hspace*{\value{ind}mm}}
\newcommand{\firstline}{\stepcounter{linenum}{\scriptsize \arabic{linenum}}\>}
\newcommand{\lref}[1]{\linenumref{#1}} % use this to refer to a line number
% End of stuff for entering source code=====================================

\newcommand{\postnotnull}{(1)}
\newcommand{\postl}{(2)}
\newcommand{\postpup}{(3)}
\newcommand{\postnonempty}{(4)}
\newcommand{\postgpnotnull}{(4a)}
\newcommand{\postp}{(4b)}
\newcommand{\postgpup}{(4c)}

\newcommand{\tabtabcom}{\>\>\com}
\newcommand{\tabcom}{\>\com}


\newcommand{\Key}{Key}
\newcommand{\R}{{\bf R}}
\newcommand{\NULL}{\mbox{$\bot$}}
\newcommand{\clean}{\mbox{\sc Clean}}
\newcommand{\mk}{\mbox{\sc Mark}}
\newcommand{\insertflag}{\mbox{I\Flag}}
\newcommand{\deleteflag}{\mbox{D\Flag}}
\newcommand{\record}{\mbox{record}}
\newcommand{\info}{\mbox{info}}
\newcommand{\FAIL}{\mbox{\sc Fail}}

\newcommand{\Flag}{Flag}
\newcommand{\DFlag}{DInfo}
\newcommand{\IFlag}{IInfo}

\newcommand{\ie}{{\it i.e.}}
\newcommand{\TRUE}{\mbox{\sc True}}
\newcommand{\FALSE}{\mbox{\sc False}}

\newcommand{\func}[1]{\mbox{\sc #1}}

\newcommand{\CASB}{\func{CAS}}

\newcommand{\comnospace}{\mbox{$\triangleright$}}
\newcommand{\com}{\mbox{\comnospace\ }}

\newcommand{\doublespace}{\addtolength{\baselineskip}{.8\baselineskip}}
\newcommand{\ds}{\addtolength{\baselineskip}{.9\baselineskip}}

%%%%%%%%%%%%%%%%%%%%% COMMENTS %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%% New state-of-the-art method for comment insertion
%%%% by multiple distributed authors.
%%%% These comments are present in the LATEX file and
%%%% do not appear in print.
%
%% define debug: use first def for debug, second def for final version
\newcommand{\debug}[1]{#1}
%\newcommand{\debug}[1]{}
%

%% \newcommand{\comment}[1]{\debug{\marginpar {\sloppy\tiny #1}}}
\newcommand{\lcomment}[1]{\debug{\comment{$\rightarrow$ #1}}}
\newcommand{\incomment}[1]{\debug{[[[#1]]]}}
%
\newcommand{\fM}[1]{\comment{#1 5-14 M.}}
\newcommand{\fD}[1]{\comment{#1 5-21 D.}}
\newcommand{\upcom}[1]{#1}
%
%%%%%%%%%%%%%%%% end of comments %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%



\newenvironment{remark}{\begin{trivlist}
		\item[\hspace{\labelsep}{\bf\noindent Remark. }]}{\end{trivlist}}
%
%

\newenvironment{centre}{\begin{center}}{\end{center}}
% sets up centre as alternate name for center.



% marginal comment
\newcommand{\comment}[1]{\marginpar{\tiny #1}}
%\mnote{Example: only one algorithm is typed in so far.}
\newcommand{\mnote}[1]
{\marginpar%
	[{\tiny\begin{minipage}[t]{\marginparwidth}\raggedright#1%
	\end{minipage}}]%
	{\tiny\begin{minipage}[t]{\marginparwidth}\raggedright#1%
	\end{minipage}}%
}



%\newcommand{\here}[1]{[[[#1]]]\marginpar{***}}
\newcommand{\ignore}[1]{}
\newcommand{\floor}[1]{\left\lfloor #1 \right\rfloor}

\newcommand{\lft}{\mbox{\it left}}




\section{Robust BST}

The original BST algorithm does not support the crash-recovery model. It is clear from the code a process does not persist the operation's response in the non-volatile memory, and thus, once a process crash the response is lost. For example, assume a process $q$ apply $\func{Insert} (k)$, performs a successful \CASB\ in line \lref{iflag-cas} and fails after completing the \func{HelpInsert} routine. In this case, the \func{Insert} operation took effect, that is, the new key appears as a leaf in the tree, and any $\func{Find} (k)$ operation will return it. However, even though the operation must be linearized before the crash, upon recovery process $q$ is unaware of it. Moreover, looking for the new leaf in the tree may be futile, as it might be $k$ has been removed from the tree after the crash.

Furthermore, if no recover routine is supplied, it may result an execution which is not well-formed. Consider for example the following scenario. A process $q$ invoke an $Op_1 = \func{Insert}(k_1)$ operation. $q$ performs a successful \CASB\ in line \lref{iflag-cas} followed by a crush. After recovering, $q$ invoke an $Op_2 = \func{Insert}(k_2)$ operation. Assume $k_1$ and $k_2$ belongs to a different parts of the tree (do not share parent or grandparent). Then, $q$ can complete the insertion of $k_2$ without having any affect on $k_1$. Now, a process $q'$ performs $\func{Find}(k_1)$ which returns \NULL, as the insertion of $k_1$ is not completed, followed by $\func{Find}(k_2)$, which returns the leaf of $k_2$. The $\func{Insert}(k_1)$ operation will be completed later by any \func{Insert} or \func{Delete} operation which needs to make changes to the flagged node. We get that $Op_2$ must be before $Op_1$ in the linearization, although $Op_1$ invoked first.

The kind of anomaly described above can be addressed by having the first \CASB\ of a successful attempt for \func{Insert} or \func{Delete} as the linearization point, as in the Linked-List. For that, the \func{Find} routine should take into consideration future unavoidable changes, for example, a node flagged with \insertflag\ ensures an insertion of some key. A simple solution is to change the \func{Find} routine such that it also helps other operations, as described in figure \ref{robust find - solution 1}. The \func{Find} routine will search for key $k$ in the tree. If the \func{Search} routine returns a grandparent or a parent that is flagged, then it might be that an insert or delete of $k$ is currently in progress, thus we first help the operation to complete, and then search for $k$ again. Otherwise, if $gpupdate$ or $pupdate$ has been changed since the last read, it means some change already took affect, and there is a need to search for $k$ again. If none of the above holds, there is a point in time where $gp$ points to $p$ which points to $l$, and there is no attempt to change this part of the tree. As a result, if $k$ is in the tree at this point, it must be in $l$, and the find can return safely. 

The approach described above is not efficient in terms of time. We would like a solution which maintain the desirable behaviour of the original \func{Find} routine, where a single \func{Search} is needed. A more refined solution is given in figure \ref{robust find - solution 2}. The intuition for it is drown from the Linked-List algorithm.
In the Linked-List algorithm it was enough to consider a marked node as if it has been deleted, without the need to complete the deletion. Nonetheless, the complex BST implementation is more challenging, as the \func{Delete} routine needs to successfully capture two nodes using \CASB\ in order to complete the deletion. Therefore, if a process $p$ executes $\func{Find}(k)$ procedure, and observes a node flagged with $\deleteflag$ attempting to delete the key $k$, it can not know whether in the future this delete attempt will succeed or fail, and thus does not know whether to consider the key $k$ as part of the tree or not. To overcome this problem, in such case the process will first try and validate the delete operation by marking the relevant node. According to whether the marking attempt was successful, the process can conclude if the delete operation is successful or not.
In order to easily implement the modified \func{Find} routine there is a need to conclude from \IFlag\ what is the new leaf (leaf $new$ in the \func{Insert} routine). For simplicity of presentation, we do not add this field, and abstractly refer to it in the code.

The correctness of the two suggested solutions relies on the following argument.
Once a process flags a node during operation $Op$ with input key $k$ (either \func{Insert} or \func{Delete}), then if this attempt to complete the operation eventually succeed (i.e., the marking is also successful in the case of \func{Delete}), then any \func{Find}(k) operation invoked from this point consider $Op$ as if it is completed.

The suggested modification, although being simple and local, only guarantee the implementation satisfy R-linearzability. However, the problem of response being lost in case of a crash is not addressed. Roughly speaking, the critical points in the code for recovery are the \CASB\ primitives, as a crash right after applying \CASB\ operation results the lost of the response, and in order to complete the operation the process needs to know the result of the \CASB. In addition, because of the helping mechanism, a suspended \func{Delete} operation which flagged a node and yet to mark one, may be completed by other process in the future, and may not. Upon recovery, the process needs to distinguish between the two cases, in order to obtain the right response.

To address this issue, we expend the helping mechanism so that it also update the info structure in case of a success. This is done by adding a boolean field, done, to the Flag structure. This way, if a process crash along an operation $Op$, upon recovery it can check to see if the operation was already completed. A crucial point is to update the $done$ bit before performing the unflagging. Therefore, if a node is no longer flagged we can be sure done was already updated.
If we switch the order, then it might be an operation and unflagging were completed, but the done bit is yet to be updated. Therefore, other processes can change the BST structure. However, if the process crash and recover at this point, the done bit is off, and the BST structure has been changed, so it will be harder for the process to conclude whether the operation took affect.

Before a process $q$ attempt to perform an operation, as it creates the \Flag\ structure $op$ describing the operation and its affect on the data structure, the process stores $op$ in a designated location in the shared memory (for simplicity, we use an array). As a result, upon recovery $q$ has an access to this information. Now, $q$ can check to see if the operation is still in progress, i.e., if the relevant node (parent or grandparent) is still flagged. If so, it first tries to complete the operation. Otherwise, it implies either the operation was completed, and therefore done bit is updated, or that the attempt was unsuccessful and there wa no write to the done bit. Hence, the done bit can distinguish between the two scenarios.
Notice that there is a scenario in which process $q$ recovers and observes an operation $Op$ as it in a progress, but just before it retries it, some other process complete the operation. We need to prove that even in such case, the operation will affect the data structure exactly once, and the right response is returned.

The given implementation does not recover the \func{Find} routine, since this routine does not make any changes to the BST, hence it is always safe to consider it as having no linearization point and reissue it. Also, for ease of presentation, we only write to $Announce[id]$ once we are about to capture a node using a CAS. However, writing to $Announce[id]$ at the beginning of the routine may be helpful in case of a crash early in the routine, so that the process will be able to use the data stored in $Announce[id]$ in such case also. The same is true with response value, $Announce[id]\rightarrow done$ is updated only if the routine made changes to the BST.

\subsubsection{Correctness}

In the following section we give a proof sketch for the algorithm correctness. We assume for simplicity nodes and \Flag\ records are always allocated new memory locations, although it is enough to require no location is reallocated as long as there is a chain of pointers leading to it. The proof relies on the correctness of the original algorithm, which can be found on [....].

The proof relies on several key arguments given below.

\newcommand{\argAnonymous}{arg1}
[\textbf{Arg1}] The original algorithm is anonymous and uniform, i.e., any number of processes can use the BST, and there is no need to know the number of processes in the system in order to use the BST. Notice that all helping routines in the given implementation are completely anonymous, and an execution of such a routine by either the process which invoked $op$ or any other helping process executes the exact same code. This observation allows the use of the following argument. If a process crash while executing some helping routine, we can consider it as an helping process which stop taking steps (more formally, there is an equivalent execution in which there is such a process, and it is indistinguishable to all process in the system). Since such process can not cause a wrong behaviour of the algorithm, so does the crash. 
A corollary of this argument is that repeating an helping routine multiple times by the same process can not violate the BST specification, as there is an equivalent executions in which multiple processes executes the different helping routines.

\newcommand{\argSearch}{arg2}
[\textbf{Arg2}] It is easy to verify the post-conditions of the \func{Search} routine still holds, as they follow directly from the routine's code, and does not rely on the structure or correctness of the BST. Also, the \func{Search} routine does not make any changes to the BST, but rather simply traverse it. Therefore \func{Find} routine, which only uses \func{Search}, does not affect any process, and in case of a failure along \func{Find} execution, reissuing it satisfies NRL.

\newcommand{\argNodeRef}{arg3}
[\textbf{Arg3}] If an internal node $nd_1$ stops pointing to a node $nd_2$ at some point of the execution, it can not point to $nd_2$ again. This attributes to the fact an \func{Insert} presents a node with two new children. Therefore, if $nd_2$ is a leaf, it can either be delete, or replaced by a new copy of an \func{Insert} operation. Otherwise, $nd_2$ is an internal node, and as such, the pointer to it by $nd_1$ can not be replaced by an \func{Insert} operation (which only allows to replacement a leaf), and therefore it can only be removed from the tree.

\newcommand{\argNodeUpdate}{arg4}
[\textbf{Arg4}] The field update of a node $nd$ can have any value only once along an execution. Any attempt to perform an operation creates a new record in the memory. If $nd\rightarrow update$ is marked, it can not be unmarked or changed. Otherwise, any attempt to flag it uses a new created record $op$. If the attempt succeed, then eventually it will be unflagged while still referring to $op$. In order to replace the value again, there must be an operation reading $nd\rightarrow update$ after it was unflagged (as any operation first help a flagged node). This operation must create a new record, and thus we can use the same argument again. As a corollary, if a process successfully flag or mark a node, there was no change to the node since the last time it read the update field of the node.

\paragraph{Proof Sketch}
Assume a process $q$ performs an operation $Op$ (either \func{Insert} or \func{Delete}). If $q$ does not crash, the algorithm is identical to the original algorithm, except for the additional write to $Announce[q]$ and $op\rightarrow done$, and thus the correctness of the original algorithm can be applied. Otherwise, $q$ crash at some point, and upon recovery it reads $op$ from $Announce[q]$. This record represent the last attempt of $q$ to complete $Op$. We split the proof based on the type of operation.

$Op = \func{Insert}$. Consider the read of $op\rightarrow p\rightarrow update$ upon recovery, and denote this value by $pupdate$. If $pupdate = \langle \insertflag, op \rangle$, this implies the iflag \CASB\ in line \lref{iflag-cas} was successful and the operation is yet to complete. It might be that \func{Insert} already took affect, that is, the new key is part of the tree, but the unflagging is yet to happen. In such case, $q$ calls \func{HelpInsert}$(op)$ in order to try and complete the operation. Considering \argAnonymous, this call can not violate the BST correctness, even if it not the first time $q$ executes it. Moreover, during \func{HelpInsert} there is a write to $op\rightarrow done$, and thus after completing the routine $q$ returns \TRUE, as required.

Else $pupdate \neq \langle \insertflag, op \rangle$. There are two scenarios to consider. Either the iflag \CASB\ of $q$ in line \lref{iflag-cas} was successful or not. If it was successful, then $p\rightarrow update = \langle \insertflag, op \rangle$ at this point. The only way to change it is to first unflag $p$. To do so, a process needs to complete an \func{HelpInsert}$(op)$ routine, and in particular must write to $op\rightarrow done$. In such case, the \func{Insert} operation was completed, and $q$ returns \TRUE. Otherwise, the \CASB\ was not successful, either because it failed, or the crash was before the \CASB. In both cases, the \func{Insert} operation will not be completed, as $op$ is not stored in $p\rightarrow update$, and thus no process has an access to it. Consequently, no process can update $op\rightarrow done$, and $q$ returns \FAIL.

$Op = \func{Delete}$. Consider the read of $op\rightarrow gp\rightarrow update$ upon recovery, and denote this value by $gpupdate$. If $gpupdate = \langle \deleteflag, op \rangle$, this implies the dflag \CASB\ of $q$ in line \lref{dflag-cas} was successful, and the operation is yet to complete. As in the \func{Insert}, it might be the operation already changed the tree. After reading $gpupdate$ $q$ invokes \func{HelpDelete}$(op)$ routine. Again, following \argAnonymous, executing this multiple times by $q$ can not violate the BST correctness.
The first process to try and mark $op\rightarrow p\rightarrow update$ during an \func{HelpDelete}$(op)$ routine is the one to determine the outcome of it. If it is successful, then $p$ is marked, and the $update$ field can not be changed. That is, any \func{HelpDelete}$(op)$ execution will obtain true in line \lref{checkmark}, and will call \func{HelpMarked}$(op)$ routine. Otherwise, the \CASB\ fails, and so $p\rightarrow update$ is no longer equal to $op\rightarrow pupdate$. By \argNodeUpdate\ it will never get this value again, and thus any marking \CASB\ during a \func{HelpDelete}$(op)$ execution will fail, and there is no call to \func{HelpMarked}$(op)$. In the first case, any \func{HelpDelete}$(op)$ routine must first complete a \func{HelpMarked}$(op)$, and thus must write to $op\rightarrow done$, while in the later case, there is no write to $op\rightarrow done$, as no \func{HelpMarked}$(op)$ is ever invoked. Therefore, in both cases, when $q$ completes \func{HelpMarked}$(op)$ it reads $op\rightarrow done$ and returns the right response.


Otherwise $gpupdate \neq \langle \deleteflag, op \rangle$, and there are two scenarios to consider. If the dflag \CASB\ of $q$ in line \lref{dflag-cas} never took affect, because it either failed, or the crash preceded it, then $op$ is never written to $gp\rightarrow update$, or to any update field. Thus, no process is aware of it, and $op\rightarrow done$ remains \FALSE, resulting $q$ returning \FAIL\ as required. Else, the \CASB\ was successful, and $gp\rightarrow update$ was flagged. The only way to change it is to first unflag it, and this in turn can be done only during an \func{HelpDelete}$(op)$ routine. In this case, it can be unflagged in either the \func{HelpMarked} routine in line \lref{dunflag-cas}, or in line \lref{backtrack-cas} of the \func{HelpDelete} routine. As mention before, the first \CASB\ in line \lref{mark-cas} of an \func{HelpDelete}$(op)$ execution determines the outcome for all \func{HelpDelete}$(op)$. If it is successful, $p\rightarrow update$ is forever marked, and all \func{HelpDelete}$(op)$ must invoke \func{HelpMarked}$(op)$. Therefore, the only option to unflag $gp\rightarrow update$ is at the end of \func{HelpMarked}$(op)$ routine, and this done only after setting $op\rightarrow done$. In such case, the \func{Delete} operation took affect, and $q$ will return \TRUE. On the other hand, if the \CASB\ was not successful, then any \func{HelpDelete}$(op)$ will fail to mark $p\rightarrow update$, and hence no \func{HelpMarked}$(op)$ is ever invoked. As a result, there is no write to $op\rightarrow done$. In such case, the \func{Delete} operation did not took affect, nor will be, and indeed $q$ will return \FAIL. 





\remove{
%%%%%%%REMOVE%%%%%%%%%%%%%%
Assume process $q$ performs an \func{Insert}$(k)$ operation. As argued in \argSearch, a crash before writing to $Announce[q]$ implies no changes has been made to the BST. Assume thus $q$ executes line \lref{insert-write-announce}, i.e., $q$ stored in $Announce[q]$ a pointer to an \IFlag\ record $op$ containing all the data of the current attempt to complete the \func{Insert} routine. Since $q$ is in the middle of a while loop, it is enough to prove that if $q$ crash before the next time it writes to $Announce[q]$, if there is such write, upon recovery it will either complete its operation with the right response, or will continue to the next write to $Announce[q]$ without having any affect on the BST. Hence, the same argument can be applied once $q$ writes to $Announce[q]$ again. Notice that if $q$ does not crash, the algorithm is identical to the original algorithm, except for the additional write to $op\rightarrow done$, and thus the original proof can 


Assume $q$ performs a successful CAS in line \lref{iflag-cas}. Then, a reference to the \IFlag\ $op$ is stored in $p\rightarrow update$, which is also flagged. Following \argNodeUpdate, $p$ was not changed since the \func{Search} routine read it, and it still points to $l$.
Starting from this point, no changes can be made to $p$, except for the change point to by $op$, as the node is flagged. Now, only the first process 

Relying on the correctness of the original algorithm, no matter how many times \func{HelpInsert}(op) will be executed, the change will occur only once. This follows from the fact that many process can observe op, and will try to complete it in the future. The core for this argument is that a node never point twice to the same node.
}







%\setcounter{linenum}{0}


\begin{figure*}[h]
	\footnotesize
	
	\begin{code}
		\func{Find}($\Key\ k$) : Leaf* \{ \nlc
		\n Internal *$gp$, *$p$\nlc
		Leaf *$l$\nlc
		Update $pupdate, gpupdate$\bl
		\nlc
		
		%     Update $pupdate,gpupdate$\ul \nlc
		while (\TRUE) \{ \nlc \n
		$\langle gp, p, l, pupdate, gpupdate \rangle := \func{Search}(k)$\nlc
		if $gpupdate.state \neq \clean$ then $\func{Help}(gpupdate)$ \nlc
		else if $pupdate.state \neq \clean$ then $\func{Help}(pupdate)$ \nlc
		else if $gp \rightarrow update = gpupdate$ and $p \rightarrow update = pupdate$ then \{ \nlc \n
		if $l \rightarrow key = k$ then return $l$ \nlc
		else return \NULL \nlc
		\p \} \nlc
		\p \} \nlc
		\p \}
	\end{code}
	
	\caption{Solution 1: R-linearizable \func{Find} routine}
	\label{robust find - solution 1}
\end{figure*}



\begin{figure*}[h]
	\footnotesize
	
	\begin{code}
		\func{Find}($\Key\ k$) : Leaf* \{ \nlc
			\n Internal *$gp$, *$p$\nlc
			Leaf *$l$\nlc
			Update $pupdate, gpupdate$\bl
			\nlc
			
			%     Update $pupdate,gpupdate$\ul \nlc
			$\langle gp, p, l, pupdate, gpupdate \rangle := \func{Search}(k)$\nlc
			if $l\rightarrow key \neq k$ then \{ \nlc
				\n if ($pupdate.state = \insertflag$ and $pupdate.info$ attempt to add key $k$) then \nlc
					\n return leaf with key $k$ from $pupdate.info$ \nlc \p
				else return \NULL \nlc \p
			\} \nlc
			if ($pupdate.state = \mk$ and $pupdate.info \rightarrow l \rightarrow key = k$) then return \NULL \nlc
			if ($gpupdate.state = \deleteflag$ and $gpupdate.info \rightarrow l \rightarrow key = k$) then \{ \nlc
				\n $op := gpupdate.info$ \nlc
				$result := \CASB(op\rightarrow p\rightarrow update, op\rightarrow pupdate, \langle \mk, op\rangle)$ \tabtabcom {\bf mark \CASB}\label{mark-cas}\nlc     
				if ($result = op\rightarrow pupdate$ or $result = \langle \mk, op\rangle$) then return \NULL \label{checkmark} \tabtabcom $op\rightarrow p$ is successfully marked \nlc \p
			\} \nlc
			return $l$ \nlc \p
		\}
	\end{code}

	\caption{Solution 2: R-linearizable \func{Find} routine}
	\label{robust find - solution 2}
\end{figure*}




\begin{figure}[bt]
	\footnotesize
	\begin{code}
		\firstline
		type Update \{ \hspace*{14mm} \com stored in one CAS word\nlc
		\n  $\{\clean,\deleteflag,\insertflag,\mk\}$ $state$ \nlc
		\Flag\ *{\it info}\nlc
		\p
		\}
		\nlc
		type Internal \{ \hspace*{13.25mm} \com subtype of Node\nlc
		\n $\Key \cup \{\infty_1,\infty_2\}$ $key$\nlc
		Update $update$\nlc
		Node *\lft, *$right$\nlc
		\p\}
		\nlc
		type Leaf \{ \hspace*{18.3mm} \com subtype of Node\nlc
		%\hspace*{4mm} \com could also store auxiliary information\nlc 
		\n $\Key \cup \{\infty_1,\infty_2\}$  $key$\nlc
		\p\}
		\nlc
		type \IFlag\ \{ \hspace*{17.7mm} \com subtype of \Flag\nlc
		\n Internal *$p$, *$newInternal$\nlc
		Leaf *$l$\nlc
		\textcolor{blue}{boolean $done$} \nlc
		\p\}
		\nlc
		type \DFlag\ \{ \hspace*{16.4mm} \com subtype of \Flag\nlc
		\n Internal *$gp$, *$p$\nlc
		Leaf *$l$\nlc
		Update $pupdate$\nlc
		\textcolor{blue}{boolean $done$} \nlc
		\p\}   
		\ul
		\com Initialization:\nlc
		shared Internal *$Root$ := pointer to new Internal node \ul
		\n with $key$ field $\infty_2$, $update$ field $\langle \clean,\NULL\rangle$, and\ul
		pointers to new Leaf nodes with keys $\infty_1$ and\ul 
		$\infty_2$, respectively, as \lft\ and $right$ fields.
		\p 
	\end{code}
	\caption{\label{code1}Type definitions and initialization.}
\end{figure}



\begin{figure*}
	\footnotesize
	
	\begin{code}
		\func{Recover}() \{ \nlc \n
			Flag $*op = Announce[id]$ \bl \nlc
			
			if $op$ of type \IFlag\ then \{ \nlc \n
				$result := op\rightarrow p\rightarrow update$ \tabtabcom Check {\bf flag}\label{iflag-cas} \nlc
				if $result = \langle \insertflag,op \rangle$ then \func{HelpInsert}$(op)$ \tabtabcom Finish the insertion\label{finish-insert}\nlc
				%if $op\rightarrow p$ points to $op\rightarrow newInternal$ or \nlc
				%\n $op\rightarrow newInternal \rightarrow update \neq \langle \clean, \NULL\rangle$ then return \TRUE \nlc
				%\p else return \FAIL \nlc
			\p \} \nlc
			if $op$ of type \DFlag\ then \{ \nlc \n
				$result := op\rightarrow gp\rightarrow update$ \tabtabcom Check {\bf flag}\label{dflag-cas}\nlc
				if $result = \langle \deleteflag, op \rangle$ then \func{HelpDelete}$(op)$ \label{delete-true} \tabtabcom Either finish deletion or unflag \nlc
				%if $op\rightarrow p\rightarrow update = \langle \mk, op \rangle$ then return \TRUE \nlc
				%else return \FAIL \nlc
			\p \} \nlc
			if $op\rightarrow done = \TRUE$ then return \TRUE\ \nlc
			else return \FAIL \nlc
		\p \}
			
			
	\end{code}

\caption{\func{Recover} routine}
\end{figure*}



\begin{figure*}
	\scriptsize
	\begin{code}
		\firstline
		\func{Search}($\Key\ k$) : $\langle \mbox{Internal*}, \mbox{Internal*}, \mbox{Leaf*}, \mbox{Update}, \mbox{Update}\rangle$  \{\ul
		\n \com Used by \func{Insert}, \func{Delete} and \func{Find} to traverse a branch of the BST; satisfies following {\it postconditions}:\ul
		\com \postnotnull\ $l$ points to a Leaf node and $p$ points to an Internal node\ul
		\com \postl\ Either $p\rightarrow \lft$ has contained $l$ (if $k<p\rightarrow key$) or $p\rightarrow right$ has contained $l$ (if $k\geq p\rightarrow key$)\ul
		\com \postpup\ $p\rightarrow update$ has contained $pupdate$\ul
		\com \postnonempty\ if $l\rightarrow key\neq \infty_1$, then the following three statements hold:\ul
		\com \hspace*{3mm}\postgpnotnull\ $gp$ points to an Internal node\ul
		\com \hspace*{3mm}\postp\ either $gp\rightarrow \lft$ has contained $p$ (if $k<gp\rightarrow key$) or $gp\rightarrow right$ has contained $p$ (if $k\geq gp\rightarrow key$)\ul
		\com \hspace*{3mm}\postgpup\ $gp\rightarrow update$ has contained $gpupdate$\nlc
		Internal *$gp$, *$p$\nlc
		Node *$l:= Root$  \label{restart-search}\nlc
		
		Update $gpupdate, pupdate$ \tabtabcom Each stores a copy of an $update$ field\bl\nlc
		%\com The initial values of $p,gp,pupdate$ and $gpupdate$ are unimportant
		%$p :=\NULL$ 
		%$pupdate := \langle \clean, \NULL\rangle$\nlc
		while $l$ points to an internal node \{ \nlc%\com Advance down the tree\nlc
		\n         $gp := p$ \tabtabcom Remember parent of $p$\nlc
		$p := l$ \tabtabcom Remember parent of $l$\nlc
		$gpupdate := pupdate$ \tabtabcom Remember $update$ field of $gp$\nlc
		$pupdate := p\rightarrow update$\label{store-pupdate}\tabtabcom Remember $update$ field of $p$\nlc  
		%           if $pupdate.state = \mk$ then \label{checkmark} \{\nlc
		%\n              \func{HelpMarked}$(pupdate.info)$ \label{call-hm3}\nlc
		%                goto line \lref{restart-search} \com Restart traversal\nlc
		%\p         \}\nlc        
		if $k < l\rightarrow key$ then $l:= p\rightarrow \lft$ else $l:=p \rightarrow right$ \label{read-child}\tabtabcom Move down to appropriate child\nlc
		%           if $pupdate \neq p\rightarrow update$ then goto line \lref{restart-search} \com Restart traversal \label{reread-pupdate}\nlc
		\p \} \nlc
		return $\langle gp, p, l, pupdate, gpupdate \rangle$ \nlc
		\p 
		\}\bl
		\nlc
		\func{Find}($\Key\ k$) : Leaf* \{ \nlc
		\n   Leaf *$l$\bl
		\nlc
		%     Update $pupdate,gpupdate$\ul \nlc
		$\langle -,-,l,-,-\rangle := \func{Search}(k)$\nlc
		if $l\rightarrow key = k$ then return $l$\nlc
		else return \NULL\nlc
		\p
		\}\bl
		\nlc
		\func{Insert}($\Key\ k$) : boolean \{ \nlc
		\n Internal *$p$, *$newInternal$\nlc 
		Leaf *$l$, *$newSibling$\nlc 
		Leaf *$new :=$ pointer to a new Leaf node whose $key$ field is $k$  \nlc
		Update $pupdate, result$\nlc
		\IFlag\ *$op$\bl\nlc%:=$ pointer to a new \IFlag\ \record\ \ul\nlc
		while \TRUE\ \{  \nlc
		\n $\langle -, p, l, pupdate, - \rangle := \func{Search}(k)$ \label{ins-search}\nlc
		if $l \rightarrow key = k$ then return \FALSE\ \tabtabcom Cannot insert duplicate key\label{insert-false}\nlc
		if $pupdate.state \neq \clean$ then \func{Help}$(pupdate)$ \tabtabcom Help the other operation \label{ins-help-unclean}\nlc
		else \{\nlc
		\n        $newSibling :=$ pointer to a new Leaf whose key is $l\rightarrow key$\nlc
		$newInternal :=$ pointer to a new Internal node with $key$ field $\max(k, l \rightarrow key)$,\label{create-internal}\ul      
		\n      $update$ field $\langle \clean, \NULL\rangle$, and with two
		child fields equal to $new$ and $newSibling$\ul 
		(the one with the smaller key is the left child)\nlc
		\p        $op :=$ pointer to a new \IFlag\ \record\  containing $\langle p, l, newInternal, \textcolor{blue}{\FALSE}\rangle$\label{new-IFlag}\nlc
		\textcolor{blue}{$Announce[id] := op$} \label{insert-write-announce} \nlc
		$result := \CASB(p\rightarrow update, pupdate, \langle \insertflag, op\rangle)$ \tabtabcom {\bf iflag \CASB}\label{iflag-cas} \nlc
		if $result = pupdate$ then \{ \tabtabcom The iflag \CASB\ was successful\nlc
		\n            \func{HelpInsert}$(op)$ \tabtabcom Finish the insertion\label{finish-insert}\nlc
		return \TRUE\ \label{insert-true}\nlc
		\p        \}\nlc 
		else \func{Help}$(result)$ \tabcom The iflag \CASB\ failed; help the operation that caused failure\label{ins-help-after-failure}\nlc
		\p    \}\nlc
		\p\}\nlc 
		\p
		\}\bl
		\nlc
		
		\func{HelpInsert}(\IFlag\ *$op$) \{\ul
		\n     \com {\it Precondition}:  $op$ points to an \IFlag\ \record\  (\ie, it is not $\NULL$)\nlc
	%	\textcolor{blue}{$op \rightarrow done := \insertflag$} \tabtabcom {announce the iflag \CASB\ was successful}\nlc 
		\func{CAS-Child}$(op\rightarrow p, op\rightarrow l, op\rightarrow newInternal)$ \tabtabcom {\bf ichild \CASB}\label{ichild-cas}\nlc
		\textcolor{blue}{$op \rightarrow done := \TRUE$} \tabtabcom {announce the operation completed}\nlc 
		$\CASB(op\rightarrow p\rightarrow update, \langle \insertflag, op \rangle, \langle \clean, op\rangle)$ \tabtabcom {\bf iunflag \CASB} \label{iunflag-cas}\nlc
		\p
		\}
	\end{code}
	\caption{\label{code2}Pseudocode for \func{Search}, \func{Find} and \func{Insert}.}
\end{figure*}

\begin{figure*}
	\scriptsize
	\begin{code}
		\firstline
		\func{Delete}($\Key\ k$) : boolean \{\nlc
		\n Internal *$gp$, *$p$\nlc
		Leaf *$l$\nlc
		Update $pupdate, gpupdate, result$\nlc
		\DFlag\ *$op$\bl\nlc
		while \TRUE\ \{ \nlc
		\n     $\langle gp, p, l, pupdate, gpupdate \rangle := \func{Search}(k)$\label{del-search}\nlc
		if $l\rightarrow key \neq k$ then return \FALSE\ \tabtabcom Key $k$ is not in the tree\label{delete-false}\nlc
		%       assert: $gp$ points to an internal node\label{assert-gpnotnull}\nlc
		if $gpupdate.state \neq \clean$ then \func{Help}$(gpupdate)$ \label{del-help-unclean-1}\nlc
		else if $pupdate.state \neq \clean$ then \func{Help}$(pupdate)$\label{del-help-unclean-2}\nlc
		else \{ \tabtabcom Try to flag $gp$\nlc
		\n          $op :=$ pointer to a new \DFlag\ \record\  containing $\langle gp, p, l, pupdate, \textcolor{blue}{\FALSE} \rangle$\label{new-DFlag}\nlc
		\textcolor{blue}{$Announce[id] := op$} \nlc
		$result := \CASB(gp\rightarrow update, gpupdate, \langle \deleteflag, op \rangle)$ \tabtabcom {\bf dflag \CASB}\label{dflag-cas}\nlc
		if $result = gpupdate$ then \{ \tabtabcom \CASB\ successful \nlc
		\n             if \func{HelpDelete}$(op)$ then return \TRUE \label{delete-true} \tabtabcom Either finish deletion or unflag\nlc
		\p          \}\nlc                 
		else \func{Help}$(result)$ \tabcom The dflag \CASB\ failed; help the operation that caused the failure \label{del-help-after-failure}\nlc%; help operation that is in the way 
		\p     \}\nlc
		\p \}\nlc
		\p
		\}\bl
		\nlc
		
		\func{HelpDelete}(\DFlag\ *$op$) : boolean \{\ul
		\n   \com {\it Precondition}:  $op$ points to a \DFlag\ \record\  (\ie, it is not \NULL)\nlc%$op\rightarrow pupdate.state = \clean$\nlc
		Update $result$ \tabtabcom Stores result of mark \CASB\bl\nlc
		%\com First, try to mark the node that $op\rightarrow p$ points to\nlc
		%$pupdate = op\rightarrow pupdate$\nlc
		$result := \CASB(op\rightarrow p\rightarrow update, op\rightarrow pupdate, \langle \mk, op\rangle)$ \tabtabcom {\bf mark \CASB}\label{mark-cas}\nlc     
		if $result = op\rightarrow pupdate$ or $result = \langle \mk, op\rangle$ then \label{checkmark}\{\tabtabcom $op\rightarrow p$ is successfully marked\nlc
		\n          \func{HelpMarked}$(op)$ \label{call-hm1} \tabtabcom Complete the deletion\nlc
		return \TRUE\tabtabcom Tell \func{Delete} routine it is done\nlc
		\p       \}\nlc
		else \{\tabtabcom The mark \CASB\ failed \nlc
		\n              
		\func{Help}$(result)$\label{help-after-failed-mark}\tabtabcom Help operation that caused failure\nlc
		\CASB$(op\rightarrow gp\rightarrow update, \langle \deleteflag, op\rangle, \langle \clean, op\rangle)$ \tabtabcom {\bf backtrack \CASB}\label{backtrack-cas}\nlc
		
		return \FALSE\tabtabcom Tell \func{Delete} routine to try again\nlc
		\p       \}\nlc
		\p
		\}\bl
		\nlc
		
		\func{HelpMarked}(\DFlag\ *$op$) \{\ul
		\n   \com {\it Precondition}:  $op$ points to a \DFlag\ \record\  (\ie, it is not $\NULL$)\nlc
		Node *$other$\bl\ul
		\com Set $other$ to point to the sibling of the node to which  $op\rightarrow l$ points \nlc
		if $op\rightarrow p\rightarrow right = op\rightarrow l$ then $other := op\rightarrow p \rightarrow \lft$ else $other:=op\rightarrow p\rightarrow right$\label{read-other}\ul 
		\com Splice the node to which $op\rightarrow p$ points out of the tree, replacing it by $other$\nlc
		\func{CAS-Child}$(op\rightarrow gp, op\rightarrow p, other)$ \tabtabcom {\bf dchild \CASB}\label{dchild-cas}\nlc
		%     \com Unflag the node $op\rightarrow gp$ points to\nlc
		\textcolor{blue}{$op \rightarrow done := \TRUE$} \tabtabcom {announce the operation completed} \nlc
		\CASB$(op\rightarrow gp\rightarrow update, \langle \deleteflag, op\rangle, \langle \clean, op\rangle)$ \tabtabcom {\bf dunflag \CASB}\label{dunflag-cas}\nlc
		\p
		\}\bl\nlc
		
		\func{Help}(Update $u$) \{ \tabtabcom General-purpose helping routine\ul
		\n    \com {\it Precondition}:  $u$ has been stored in the $update$ field of some internal node\nlc
		if $u.state = \insertflag$ then \func{HelpInsert}$(u.\info)$\label{call-HelpInsert}\nlc
		else if $u.state = \mk$ then \func{HelpMarked}$(u.\info)$\label{call-hm2}\nlc
		else if $u.state = \deleteflag$ then \func{HelpDelete}$(u.\info)$\label{call-HelpDelete}\nlc
		\p
		\}\bl
		\nlc
		\func{CAS-Child}(Internal *$parent$, Node *$old$, Node *$new$) \{\label{CAS-Child}\ul
		\n  \com {\it Precondition}:  $parent$ points to an Internal node and $new$ points to a Node (\ie, neither is \NULL)\ul
		\com This routine tries to change one of the child fields of the node that $parent$ points to from $old$ to $new$.\nlc
		if $new \rightarrow key < parent\rightarrow key$ then\label{which-child}\nlc
		\n       \CASB$(parent\rightarrow \lft, old, new)$\label{child-cas-1}\nlc
		\p  else\nlc
		\n       \CASB$(parent\rightarrow right, old, new)$\label{child-cas-2}\nlc
		\p\p 
		\}
	\end{code}
	\caption{\label{code3}Pseudocode for \func{Delete} and some auxiliary routines.}
\end{figure*}